---
title: "p8105_hw2_zh2596"
author: "Zilin Huang"
date: "`r Sys.Date()`"
output: github_document
---

## Problem 1

```{r message=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
```

```{r, message=FALSE, echo=FALSE}
unemployment = read_csv("data/unemployment.csv")
snp = read_csv("data/snp.csv")
```

First, clean the data in `pols-month.csv`.

```{r, message=FALSE}
pols_month = 
  read_csv("data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year","month","day"), sep = "-") |>
  mutate(month = month.abb[as.numeric(month)],
         president = ifelse(prez_gop==1,"gop","dem"))
  
pols_month = select(pols_month, -day, -prez_gop, -prez_dem)
  
```

Second, clean the data in `snp.csv` using a similar process to the above.

```{r message=FALSE}
snp = 
  read_csv("data/snp.csv") |>
  janitor::clean_names() |>  
  mutate(
    date = as.Date(date, format = "%m/%d/%y"),
    date = as.Date(ifelse(date > Sys.Date(), format(date,"19%y-%m-%d"),
                                                format(date)))) |>
  separate(date, into = c("year","month","day"), sep = "-") |>
  mutate(month = month(as.numeric(month), label=TRUE, abbr=TRUE),
         year=as.character(year))|>
  select(-day) |>
  arrange(year,month)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets.

```{r message=FALSE}
unemployment =
  read_csv("data/unemployment.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment rate"
  ) |>
  mutate(month = str_to_title(month),
         year = as.character(year))
```

Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.

```{r message=FALSE, warning=FALSE}
# merge `snp` into `pols_month`:
merge_df =
  left_join(pols_month, snp, by = c("year","month"))

# merging `unemployment`:
merge_df = 
  left_join(merge_df, unemployment, by = c("year","month"))
```

`pols_month` contains the summary of politicians who are democratic or republican, as well as the party that the president belongs to during each month, from Jan. 1947 to Jun. 2015;
`snp` contains the the closing values of the S&P stock index per month from Jan. 1950 to July. 2015;
`unemployment` contains the resulting dataset contains the unemployment rate as percentage value per month from Jan. 1948 to Jun. 2015;

The resulting dataset contains `r nrow(merge_df)` rows and `r ncol(merge_df)` columns, and the range of year is `r as.numeric(max(merge_df$year)) - as.numeric(min(merge_df$year))`. Some of the key variables are the party of the president (`president`), the closing values of the S&P stock index (`close`), and the `unemployment rate` during a specific month.

## Problem 2

Read and clean the Mr. Trash Wheel sheets, and compute a new column called `homes_powered`.

```{r message=FALSE}
Mr_Trash_Wheel = 
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
                     sheet = "Mr. Trash Wheel",
                     range = "A2:N549") |>
  mutate(homes_powered = floor(`Weight (tons)` * 500 / 30), name = "Mr.Trash", 
         Year = as.character(Year)) |>
  arrange(Year, Month) |>
  relocate(name, Year, Month)
```

```{r message=FALSE}
Professor_Trash_Wheel = 
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
                     sheet = "Professor Trash Wheel",
                     range = "A2:M96") |>
  mutate(homes_powered = floor(`Weight (tons)` * 500 / 30), name = "Professor",
         Year = as.character(Year)) |>
  arrange(Year, Month) |>
  mutate(name = "Professor") |>
  relocate(name, Year, Month)
```

```{r message=FALSE}
Gwynnda_Trash_Wheel = 
  readxl::read_excel("data/202309 Trash Wheel Collection Data.xlsx", 
                     sheet = "Gwynnda Trash Wheel",
                     range = "A2:K108") |>
  mutate(homes_powered = floor(`Weight (tons)` * 500 / 30), name = "Gwynnda",
         Year = as.character(Year)) |>
  arrange(Year, Month) |>
  relocate(name, Year, Month)
```

Combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset.

```{r message=FALSE, warning=FALSE}
merge_df_2 =
  left_join(Professor_Trash_Wheel, Gwynnda_Trash_Wheel, by = c("Year", "Month"))

merge_df_2 = 
  left_join(Mr_Trash_Wheel, merge_df_2, by = c("Year", "Month"))
```

There are `r nrow(merge_df_2)` observations in this dataset. Some of the key variables include the dumpster ID, Weight and Volume of trash dumped for each of the three vessels. 

```{r}
# Answers to the following two questions:
total_weight=sum(Professor_Trash_Wheel$`Weight (tons)`)
total_butt=sum(filter(Gwynnda_Trash_Wheel,Year=="2021",Month=="July")$`Cigarette Butts`)
```

### What was the total weight of trash collected by Professor Trash Wheel?

The total weight is `r total_weight` tons.

### What was the total number of cigarette butts collected by Gwynnda in July of 2021?

The total number is `r total_butt`.

## Problem 3

Import, clean, and tidy the dataset of baseline demographic.

```{r message=FALSE}
# This step imports the csv file of baseline demographic, and 
# ensure that sex and APOE4 carrier status are appropriate encoded:
baseline = 
  read_csv("data/MCI_baseline.csv", skip=1) |>
  mutate(Sex = 
           case_match(Sex,
                       1 ~ "Male",
                       0 ~ "Female"),
         apoe4 = 
           case_match(apoe4,
                      1 ~ "carrier",
                      0 ~ "non-carrier")) |>
  # This step filters out the rows containing null values for "Age at onset",
  # as well as rows in which the value of "Current age" 
  # is larger than that of "Age at onset".
  filter(!(`Current Age` > `Age at onset` | is.na(`Age at onset`)))

```

The MCI_baseline csv file includes each participant's current age at the beginning of the study, demographic information including sex and education level, as well as records of whether carrying APOE4 and age at onset of MCI.

```{r}
# Answers to the following three questions:
MCI_participants = nrow(filter(baseline, `Age at onset` >= 0))
avg_baseline_age = mean(baseline$`Current Age`)
```

### How many participants were recruited, and of these how many develop MCI? 

There are `r nrow(read_csv("data/MCI_baseline.csv", skip=1))` participants recruited, and `r MCI_participants` participants develop MCI.

### What is the average baseline age?

The average baseline age is `r round(avg_baseline_age, 2)`.

### What proportion of women in the study are APOE4 carriers?

The proportion is `r round(nrow(filter(baseline,Sex == "Female",apoe4 == "carrier"))/nrow(filter(baseline,Sex=="Female"))*100, 2)`%.

Import, clean, and tidy the dataset containing longitudinal biomarker values.

```{r, message=FALSE}
# This step imports the csv file of amyloid.
amyloid = 
  read_csv("data/mci_amyloid.csv", skip = 1) |>
  rename(ID = `Study ID`) |>
  # This step stacks the values of different time periods into one column.
  pivot_longer(
    `Time 2`: `Time 8`,
    names_to = "time period", 
    values_to = "amyloid values")
```

The mci_amyloid csv file contains each participant's amyloid $\beta$ 42/40 ratio that changes over the study period.

Combine these two datasets so that only participants who appear in both datasets are retained.

```{r}
# Combine the datasets:
merge_df_3 = 
  inner_join(baseline, amyloid, by = "ID")
```

This dataset consists of the recorded information, including age at MCI's onset, amyloid levels at different time slots, etc., of participants from both the baseline and the amyloid datasets.

Next, export it as a CSV to the data directory.

```{r}
write.csv(merge_df_3, "data/Problem_3_Dataset.csv")
```



